Augmentation of a Historical Harpsichord Keyboard Replica for Haptic-Enabled Interaction in Museum Exhibitions

### 2025-01-31 src/article.tex

\begin{figure}
    \centering    
    \includegraphics[width=\linewidth]{src/images/3-key-side.png}
    \caption{3-Key Model Harpsichord Mechanism acquired from San Colombano, Bologna, 2023}
    \label{fig:3key}
\end{figure}

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{src/images/tagging-jacks-3.jpg} 
  \caption{Gradient stickers applied to the side of the jack body. The coarse gradient scale was selected to maximise signal excursion while preserving signal readout stability.}
  \Description{} 
  \label{fig:jack-tags}
\end{figure}

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{src/images/simple-schematic-bw-.jpg} 
  \caption{Optical sensor in a simple voltage divider circuit. \texttt{V\_OUT} is routed to one of 8 channels on the CD4051BE multiplexer}
  \Description{} 
  \label{fig:simple-schematic}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/sensor-board-w-baffles.jpeg}
    \\
    \includegraphics[width=\linewidth]{src/images/baffles.png}
    \caption{Baffles designed to prevent cross-talk between adjacent sensors.}
    \Description{}
    \label{fig:baffles}
\end{figure}
\subsection{Controller Board}\label{controller-board}

@@

The design of the augmented replica keyboard for the Tagliavini Collection was guided by three principal constraints stipulated by the museum:
\begin{itemize}
\item \emph{No visible electronics}. To preserve the visual integrity of the exhibit, all electronic components had to remain concealed except for a pair of headphones and a small display for audio parameter adjustments.
\item \emph{Robustness and reliability}. The system needed to accommodate frequent use by museum visitors and allow for straightforward maintenance by staff without requiring specialised technical expertise.
\item \emph{Faithfulness}. The keyboard mechanism ensured fidelity to its authentic operation, keeping artefacts to a minimum. 
\end{itemize}
These constraints demanded a non-invasive, robust design that was respectful of the historical authenticity of the instruments. The requirement for faithfulness to the original mechanism immediately ruled out a mechanical design reliant on electromechanical actuators, as in previous works on piano haptics \cite{Timmermans2020,Gillespie1996}. While actuators can generate considerable force, they cannot replicate the wideband signals characteristic of the harpsichord's sharp and responsive haptics (A REFERENCE WOULD BE GOOD HERE, ANY IDEAS?), which differ significantly from a piano.

%\subsection{Summary of the Finalised Design}

The final design, visible in Figures \ref{fig:teaser} and \ref{fig:details}, is a 49-key harpsichord keyboard replica equipped with an optical sensor, whose voltage output depends on the amount of light reflected by sensor surface. In this case, a sensor surface is a grey-scale gradient printed on a vinyl sticker and applied to the side of each jack (Figure \ref{fig:jack-tags}).

\subsubsection{Alternative}
\textcolor{red}{
The final design, visible in Figures \ref{fig:teaser} and \ref{fig:details}, is a 49-key two-register harpsichord keyboard replica.
The keyboard was consciously chosen to mimic that of the original XVI century Italian harpsichords. This design decision was made in order to use the propensity humans have to being influenced by visual components when musical making judgements \cite{Tsay2013} to our advantage. It is especially true with respect to musical instruments as the studies by Fritz \emph{et al}\cite{Fritz2012, Fritz2014, Fritz2017} have clearly shown. Rather than avoiding this effect we have chosen instead to utilise it. This `musical instrument McGurk effect' is in part the reason behind hiding the electronics out of sight. 
The aesthetics of the interface make it more likely that the interface be accepted on the same terms as an `authentic' musical instrument. With all the work into modelling and recreating piano action \cite{Cadoz1990, Gillespie1996, Timmermans2020} it is possible the visual component my provide the necessary push of acceptance similar to that which influences our judgment on performance \cite{Tsay2013}. 
% Such an effect is invaluable if visitors are expected to engage with the interface on an equal footing as the historic instruments that neighbour it in the museum.
% The keyboard in not merely a facsimile of a harpsichord, but rather by being equipped with optical sensors, it can provide modes of expression with a digital instrument which would otherwise be inaccessible to its analogue sibling. 
The optical sensors are wired in a voltage divider configuration (Figure \ref{fig:simple-schematic}) whose voltage output depends on the amount of light reflected by sensor surface. In this case, a sensor surface is a grey-scale gradient printed on a vinyl sticker and applied to the side of each jack (Figure \ref{fig:jack-tags}). Though the current configuration simply sends MIDI note on and off at passing a threshold, the displacement of all jacks is available continuously.
The staggering of pluck point of the jack means that there are two very distinct tactile `anchors' along the key dip. A combination of string tension, quill voicing and jack staggering contribute to the overall `weight' of the action, with each key differing from the low to the high octaves \cite{Veroli2012}. Such tactile expectations provide fertile opportunities for extending the expressive capabilities of the interface. The current sensor system should for recording of traditional use and identify where those opportunities for new expression lie \cite{McPherson2013-2}. 
}

@@

\subsection{Project Deployment}
A further objective, set forth by the authors, was to ensure accessibility and reproducibility by committing to an open source approach for all outputs of the project. The commitment to open sourcing encompassed all aspects of the system, including hardware schematics, firmware, and calibration data. Cost-effectiveness was also a central consideration. The system was initially developed for a 3-key prototype, shown in Figure \ref{fig:3key}, and successfully scaled to 49 keys without significant increases in cost or complexity. Specifically, the system was designed to be easily assembled using resources typically available in a university-managed maker space.

@@

@@

The rectangular poplar frame also deviates from the traditional logarithmic form to allow for the installation of the electronic sensors, as visible in Figure \ref{fig:49-key-bottom}, without compromising the visual or tactile authenticity. Two string choirs, crafted from yellow brass wire and anchored with wrought iron pins, were tensioned to replicate authentic plucking resistance. Felt strips were added to dampen vibrations. The result is an interface that combines the mechanical action of keys with synthetic sound generation, preserving a real harpsichord's tactile qualities.

@@

%\subsection{Sensor Criteria}\label{sensor-criteria}

@@

\subsection{Sensor Board Design}\label{sensor-board}

@@

The optical sensors detect infrared light reflected from nearby surfaces, as per Figure \ref{fig:simple-schematic}. The gradient stickers affixed to each jack provided a consistent reflective surface for the sensors (Figure \ref{fig:jack-tags}), which were used to track jack displacement precisely.

3D-printed baffles were installed on the PCBs to eliminate cross-talk between adjacent sensors, as per Figure \ref{fig:baffles}. These baffles, fabricated from dark-pigmented PLA, ensured that infrared reflections from neighbouring jacks did not interfere with sensor readings. Testing confirmed that this measure significantly improved data reliability.

@@

The Arduino Nano was selected for its compact form factor and native USB support. The 33 BLE variationof the Nano had the additional benefit in its ability to be programmed as BLE MIDI device.  Initial iterations of the controller board used 2.54 mm pitch headers for cable connections, but issues with cable disconnections led to the adoption of JST-PH connectors in later designs. Custom cable looms were constructed using adapted crimping tools, ensuring robust connections without compromising accessibility.

Non-volatile memory, implemented using Ferroelectric RAM (FRAM), provided a reliable means of storing calibration data. This ensured that sensor thresholds and other settings could be preserved across power cycles, enhancing the system's usability in museums.

@@

Each sensor's threshold was manually adjusted using RGB LEDs integrated into the system. These LEDs provided visual feedback during calibration, enabling the identification of malfunctioning sensors and simplifying the alignment process. The calibration workflow utilised the Arduino IDE serial plotter and open source MIDI monitoring software, as per Figure \ref{fig:serial_monitor}. As an expert harpsichord performer, finer calibration was carried out with the guidance of \anon{Catalina Vicens}.

Power requirements for the system were estimated at 1.1 A at 5 V, with fluctuations during startup. While the sensors were powered continuously in this iteration, future designs may incorporate power-saving measures, such as dynamic activation of sensors based on usage.

@@

\subsection{Final Assembly and Integration}

The final 49-key harpsichord replica incorporated the sensor and controller boards into the rectangular frame. Ribbon cables connected the PCBs, while felt strips were interwoven between the strings to suppress vibrations and maintain the tactile response of the keys. The assembly process at the \anon{NEMUS} lab in \anon{Bologna} involved meticulous adjustments to key clearances and wiring layouts. Modifications included milling sections of the keys to ensure adequate clearance for the PCBs and varnishing the jacks to improve sticker adhesion. The sample library, purchased from Spitfire Audio and controlled via Native Instrument's Kontakt, was installed on a Mac Mini hosted inside the instrument's case. Special holes drilled at the back of the instrument's frame, away from sight, allow the passage of cables, such as from power supplies, USB connectors and headphone jack. An iPad is used as the secondary monitor, displaying the library alone. Users interact with the library through it, while the main 12-inch display sits inside a custom-made drawer locked during usage.

@@

In an effort to make the exhibition more accessible and authentic, the \anon{San Colombano Museum} commissioned the construction of a mechanical keyboard to trigger a MIDI-controlled sample library, designed for use by museum visitors. The exhibition is set to open approximately a month after this work's submission, and only preliminary feedback has been collected from a pool of testers. The keyboard shows significant promise in enhancing the museum experience. Testers unfamiliar with harpsichords were particularly impressed by its responsiveness. The keyboard is hosted in the \emph{Oratory} above the museum's main hall, an exceptional testimony to the Bolognese art, decorated by the finest students of the Carracci and presenting a series of frescoes covering most of the walls and ceiling. The room hosts unique examples of the Italian Renaissance building tradition, including the 1547 harpsichord and the 1540 spinet by Alessandro Trasuntino. Visitors often visit the room with the same caution and respect that is typical of worship spaces. Testers did not find that the exposed headphones and touchscreen affected such an experience negatively, though a more definitive answer will come as more feedback is collected after launch. Pictures of the keyboard hosted in the \emph{Oratory} are shown in Figure \ref{fig:oratory}.

Expert feedback has highlighted certain limitations, including imprecise key calibration, which creates a temporal disconnect between the tactile plucking sensation and sound onset. Additionally, the commercial sample library, while allowing the selection of registers such as 8', 4', and their combinations, restricts functionality to a single MIDI message per key, regardless of the number of string choirs controlled. Inspired by early Italian instruments, the keyboard's layout is designed to manage two 8' registers. However, one register was disengaged, and sensors were placed before a single set of jacks to accommodate these limitations. Technical improvements, such as refined calibration with hysteresis and developing a custom sample library, could address these constraints in future iterations.

A longer discussion, but one going beyond the scope of this work, is whether the current setup or its future iterations may be effectively used to build legitimate replicas of historical musical instruments or even become a kind of \emph{new} musical instrument altogether. As such, the question is whether these designs may result in music being practised, performed and recorded with the instruments. This work supports an overarching narrative extending beyond its application in museum collections. The \anon{NEMUS project \cite{NEMUS}}, focusing on developing advanced physical models simulating the non-linear interaction between subcomponents, has commissioned a second keyboard to further explore the role of control interfaces in performance, exploiting the inherently bi-directional signal paths linking the mechanical world with the virtual, replicating the embodied relationships between performer and instrument and offering opportunities to modify or even disrupt these interactions. The upcoming Rem@ke project \cite{remake1} has also expressed an interest in engaging with the keyboard to explore meaningful embodied interactions between players and instruments.

\section{Conclusion}\label{conclusion}

This study has presented the design and implementation of an electronically augmented keyboard for the Tagliavini Collection at the San Colombano Museum. By leveraging optical sensors and MIDI-controlled sample libraries, the keyboard allows users to explore a harpsichord's sonic and mechanical essence without compromising the integrity of the museum's historical artefacts. Preliminary feedback shows that the keyboard has the potential to enhance the museum experience, particularly for those unfamiliar with early keyboard instruments. However, technical limitations, such as calibration inconsistencies and restrictions imposed by the commercial sample library, highlight opportunities for refinement. These issues will be addressed in future iterations, including a second keyboard designed to control advanced physical models. This work opens new possibilities for engaging with heritage collections while respecting their authenticity. It also serves as a foundation for further research projects exploring embodied cognition in historical instruments, and it may provide a useful framework to investigate the links between physical modelling and digital musical interface design.

>>>>

@@

The design of the augmented replica keyboard for the Tagliavini Collection was guided by three principal constraints stipulated by the museum:
\begin{itemize}
\item \emph{No visible electronics}. To preserve the visual integrity of the exhibit, all electronic components had to remain concealed except for a pair of headphones and a small display for audio parameter adjustments.
\item \emph{Robustness and reliability}. The system needed to accommodate frequent use by museum visitors and allow for straightforward maintenance by staff without requiring specialised technical expertise.
\item \emph{Faithfulness}. The keyboard mechanism ensured fidelity to its authentic operation, keeping artefacts to a minimum. 
\end{itemize}

These constraints demanded a non-invasive, robust design that was respectful of the historical authenticity of the instruments. The requirement for faithfulness to the original mechanism immediately ruled out a mechanical design reliant on electromechanical actuators, as in previous works on piano haptics \cite{Timmermans2020,Gillespie1996}. 
While actuators can generate considerable force, it is difficult to achieve the free-motion and resistance of a jack without extensive mechanical intervention such as that carried out by Gillespie \cite{Gillespie1996}. A tactile realism of the harpsichord can already be achieved by the instrument itself. It should first be confirmed if mechanical recreation is sufficient as part of a digital musical instrument before elaborate electromechanical emulation.
% (A REFERENCE WOULD BE GOOD HERE, ANY IDEAS?)
, which differ significantly from a piano.

The final design, visible in Figures \ref{fig:teaser} and \ref{fig:details}, is a 49-key two-register harpsichord keyboard replica.
The keyboard was consciously chosen to mimic that of the original XVI century Italian harpsichords. This design decision was made in order to use the propensity humans have to being influenced by visual components when musical making judgements \cite{Tsay2013} to our advantage. It is especially true with respect to musical instruments as the studies by Fritz \emph{et al}\cite{Fritz2012, Fritz2014, Fritz2017} have clearly shown. Rather than avoiding this effect we have chosen instead to utilise it. This `musical instrument McGurk effect' is in part the reason behind hiding the electronics out of sight. 
The aesthetics of the interface make it more likely that the interface be accepted on the same terms as an `authentic' musical instrument. With all the work into modelling and recreating piano action \cite{Cadoz1990, Gillespie1996, Timmermans2020} it is possible the visual component may provide the necessary `push' of acceptance similar to that which influences our judgment on performance \cite{Tsay2013}.

\begin{figure}[!b]  
  \centering
  \includegraphics[width=0.9\linewidth]{src/images/tagging-jacks-3.jpg} 
  \caption{Gradient stickers applied to the side of the jack body. The coarse gradient scale was selected to maximise signal excursion while preserving signal readout stability.}
  \Description{} 
  \label{fig:jack-tags}
\end{figure}

Optical sensors are positioned in front of each jack and their output voltage depends on the amount of light reflected by a sensor surface. In this case, the sensor surface is a grey-scale gradient printed on a vinyl sticker \ref{fig:jack-tags} and applied to the side of each jack. Though the current configuration simply sends MIDI note on and off at passing a threshold, the displacement of all jacks is available continuously.
The staggering of pluck point of the jack means that there are two very distinct tactile `anchors' along the key dip. A combination of string tension, quill voicing and jack staggering contribute to the overall `weight' of the action, with each key differing from the low to the high octaves \cite{Veroli2012}. Such tactile expectations provide fertile opportunities for extending the expressive capabilities of the interface. The current sensor system enables the data recording of traditional use from which it can be identified where those opportunities for new expression lie \cite{McPherson2013-2}.

@@

\subsection{Project Deployment}

A further objective, set forth by the authors, was to ensure accessibility and reproducibility by committing to an open source approach for all outputs of the project. The commitment to open sourcing encompassed all aspects of the system, including hardware schematics, firmware, and calibration data. Cost-effectiveness was also a central consideration. The system was initially developed for a 3-key prototype, shown in Figure \ref{fig:3key}, and successfully scaled to 49 keys without significant increases in cost or complexity. Specifically, the system was designed to be easily assembled using resources typically available in a university-managed maker space.

@@

\begin{figure}[!b]
    \centering    
    \includegraphics[width=\linewidth]{src/images/3-key-side.png}
    \caption{3-Key Model Harpsichord Mechanism acquired from San Colombano, Bologna, 2023}
    \label{fig:3key}
\end{figure}

@@

% % Option 2
\begin{figure}
    \centering
     \begin{subfigure}[h]{0.4\linewidth}
        \centering
        \includesvg[inkscapelatex=false, width = \linewidth]{images/log_harp_outline}            
        \caption{logarithmic shape}    
        \label{fig:log-harp}
    \end{subfigure}

    \begin{subfigure}[h]{0.4\linewidth}
        \centering
        \includesvg[inkscapelatex=false, width = \linewidth]{images/49-key_outline}        
        \caption{rectangular shape}
        \label{fig:rect-harp}
    \end{subfigure}
    \caption{Logarithmic shape of an original Trasuntino harpsichord contrast with rectangular shape of the replica}    
    \Description{}
    \label{fig:log-harp-comp}
\end{figure}

The rectangular poplar frame \ref{fig:rect-harp} also deviates from the traditional logarithmic \ref{fig:log-harp} form to allow for the installation of the electronic sensors, as visible in Figure \ref{fig:49-key-bottom}, without compromising the visual or tactile authenticity. Two string choirs, crafted from yellow brass wire and anchored with wrought iron pins, were tensioned to replicate authentic plucking resistance. Felt strips were added to dampen vibrations. The result is an interface that combines the mechanical action of keys with synthetic sound generation, preserving a real harpsichord's tactile qualities.

@@

@@

\subsection{Sensor Board}\label{sensor-board}

\begin{figure}[!b] 
  \centering
  \includegraphics[width=0.7\linewidth]{src/images/simple-schematic-bw-.jpg} 
  \caption{Optical sensor in a simple voltage divider circuit. \texttt{V\_OUT} is routed to one of 8 channels on the CD4051BE multiplexer}
  \Description{} 
  \label{fig:simple-schematic}
\end{figure}

@@

The optical sensors are wired in a voltage divider configuration \ref{fig:simple-schematic} and the circuit outputs a voltage based on the infrared light reflected from nearby surfaces. 
The gradient stickers affixed to each jack provided a surface with varying reflectivity for the sensors, which were used to track jack displacement throughout the key dip.

3D-printed baffles were installed on the PCBs to eliminate cross-talk between adjacent sensors, as per Figure \ref{fig:baffles}. These baffles, fabricated from dark-pigmented PLA, ensured that infrared reflections from neighbouring jacks did not interfere with sensor readings. A manual visual check of plotted all sensors readings confirmed that interference from neighbouring jacks had been eliminated.

\subsection{Controller Board}\label{controller-board}

@@

The Arduino Nano small form factor made fitting the board inside the harpsichord easier and also allowed for testing of multiple chipsets. The 33 BLE variation of the Nano had the additional benefit in its ability to be programmed as BLE MIDI device.  
%Initial iterations of the controller board used 2.54 mm pitch headers for cable connections, but issues with cable disconnections led to the adoption of JST-PH connectors in later designs. Custom cable looms were constructed using adapted crimping tools, ensuring robust connections without compromising accessibility.

Non-volatile memory, implemented using a Fujitsu MB85RS64 Ferroelectric RAM (FRAM) chip, provided a reliable means of storing calibration data. This ensured that sensor thresholds and other settings could be preserved across power cycles, enhancing the system's usability in museums.

@@

A rotary encoder was used as the interface to select a key, edit a threshold and save current thresholds to the FRAM.
Thresholds were first set the midrange of possible values a a default. A rough calibration process was carried out where each sensor were selected individually and its readings were plotted against their current threshold value using the Arduino IDE serial plotter. Thresholds were adjusted with the rotary encoder until the key no longer passed the threshold line until the jack had plucked the string. During calibration RGB LEDs were used as a visual aid to quickly identify which key was currently being adjusted. These LEDs also provided visual feedback during calibration, and would change colour based on if readings were above below the threshold or if the were outside previously recorded maximum or minimum values. Such use of the LEDs meant for easy identification of malfunctioning sensors and simplified the alignment process. A finer calibration process was carried out with the guidance of expert harpsichord performer \anon{Catalina Vicens}.

Power requirements for the system were estimated at 1.1 A at 5 V, with some fluctuation when the system was first powered on. While the sensors were powered continuously in this iteration, future designs may incorporate power-saving measures, such as dynamic activation of sensors based found in the McPherson piano \cite{McPherson2013}.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth]{src/images/mac-mini.jpg}
    \caption{Components and cabling hidden in the bottom section of the instrument case.}
    \label{fig:mac-mini}
\end{figure}

% \subsection{Final Assembly and Integration}

% The final 49-key harpsichord replica incorporated the sensor and controller boards into the rectangular frame. Ribbon cables connected the PCBs, while felt strips were interwoven between the strings to suppress vibrations and maintain the tactile response of the keys. The assembly process at the \anon{NEMUS} lab in \anon{Bologna} involved meticulous adjustments to key clearances and wiring layouts. Modifications included milling sections of the keys to ensure adequate clearance for the PCBs and varnishing the jacks to improve sticker adhesion.

For the initial version of the exhibition, a harpsichord sample library was used and controlled via Native Instrument's Kontakt. The software was installed on a Mac Mini hosted inside the instrument's case. Holes were drilled at the back of the instrument's frame, away from sight, allow the passage of cables, such as from power supplies, USB connectors and headphone jack \ref{fig:mac-mini}. These components are accessible via locked drawer which is visible in \ref{fig:oratory}. An iPad is used as a monitor through which visitors can adjust parameters of playback such as tuning, voicing and which stops are engaged. The next iteration will use a library of samples from instruments in the collection, which is currently being carried out by the museum.
%A main 12-inch display sits inside a custom-made drawer locked during usage.

\begin{figure}[!hb]
    \centering
    \includegraphics[width=0.8\linewidth]{src/images/sensor-board-w-baffles.jpeg}
    \\
    \includegraphics[width=0.8\linewidth]{src/images/baffles.png}
    \caption{Baffles designed to prevent cross-talk between adjacent sensors.}
    \Description{}
    \label{fig:baffles}
\end{figure}
\subsection{Controller Board}\label{controller-board}

@@

% In an effort to make the exhibition more accessible and authentic, the \anon{San Colombano Museum} commissioned the construction of a mechanical keyboard to trigger a MIDI-controlled sample library, designed for use by museum visitors. 
The exhibition is set to open approximately a month after this work's submission, and only preliminary feedback has been collected from a pool of 20 visitors -- who were given training on the system -- and expert feedback from \anon{Catalina Vicens} and \anon{Roberto Livi}.
Collection of visitor feedback was carried out as short informal interviews and primarily as a means of identifying any technical problems.
The keyboard shows promise in enhancing the museum experience as comments about a ``sense of disconnect'' as found in the McAlpine exhibition \cite{McAlpine2014} were absent in this initial feedback.

A more formal surveying of museum visitors is to be carried out to determine to what degree the exhibition achieved it's core aims.
% Testers unfamiliar with harpsichords were particularly impressed by its responsiveness.
The keyboard is hosted in the \emph{Oratory} above the museum's main hall, an exceptional testimony to the Bolognese art, decorated by the finest students of the Carracci and presenting a series of frescoes covering most of the walls and ceiling. The room hosts unique examples of the Italian Renaissance building tradition, including the 1547 harpsichord and the 1540 spinet by Alessandro Trasuntino. Visitors often visit the room with the same caution and respect that is typical of worship spaces. Among the visitors interviewed, there were no comments made that the exposed headphones and touchscreen affected such an experience negatively, though a more definitive answer will come as more feedback is collected after launch. Pictures of the keyboard hosted in the \emph{Oratory} are shown in Figure \ref{fig:oratory}.

Expert feedback has highlighted certain limitations, including imprecise key calibration, which creates a temporal disconnect between the tactile plucking sensation and sound onset. Additionally, the commercial sample library, while allowing the selection of registers such as 8', 4', and their combinations, restricts functionality to a single MIDI message per key, regardless of the number of string choirs controlled. Inspired by early Italian instruments, the keyboard's layout is designed to manage two 8' registers with a sensor system fitted to each. Internally the instrument is therefore 2 individual MIDI devices however, one register needed to be disengaged as the software could only communicate with a single device. 
Multiple instances of the software could be used, but an arrangement of 2 sets GUI controls was deemed too confusing to visitors who would interact with the exhibition without orientation.
Sensors are therefore active only on a single set of jacks to accommodate these limitations. Technical improvements, such as refined calibration with hysteresis and developing a custom sample library and interface, could address these constraints in future iterations.

A longer discussion, but one going beyond the scope of this work, is whether the current setup or its future iterations may be effectively used to build legitimate replicas of historical musical instruments or even become a kind of \emph{new} musical instrument altogether. As such, the question is whether these designs may result in music being practised, performed and recorded with the instruments. This work supports an overarching narrative extending beyond its application in museum collections. The \anon{NEMUS project \cite{NEMUS}}, focusing on developing advanced physical models simulating the non-linear interaction between subcomponents, has commissioned a second keyboard to further explore the role of control interfaces in performance, employing the sense of touch as link between the mechanical world with the digital, replicating the embodied relationships between performer and instrument and offering opportunities to modify or even disrupt these interactions. The upcoming Rem@ke project \cite{remake1} has also expressed an interest in engaging with the keyboard to explore meaningful embodied interactions between players and instruments.

% \section{Conclusion}\label{conclusion}

% This study has presented the design and implementation of an electronically augmented keyboard for the Tagliavini Collection at the San Colombano Museum. By leveraging optical sensors and MIDI-controlled sample libraries, the keyboard allows users to explore a harpsichord's sonic and mechanical essence without compromising the integrity of the museum's historical artefacts. Preliminary feedback shows that the keyboard has the potential to enhance the museum experience, particularly for those unfamiliar with early keyboard instruments. However, technical limitations, such as calibration inconsistencies and restrictions imposed by the commercial sample library, highlight opportunities for refinement. These issues will be addressed in future iterations, including a second keyboard designed to control advanced physical models. This work opens new possibilities for engaging with heritage collections while respecting their authenticity. It also serves as a foundation for further research projects exploring embodied cognition in historical instruments, and it may provide a useful framework to investigate the links between physical modelling and digital musical interface design.
